Copulas and their potential for ecology

### 2019-06-13 Paper.Rmd
done with round 1 of prepping for Ecol Mongr, going through it one more time now

\noindent \emph{Type of Submission:} Article

@@

All branches of ecology study relationships among environmental and biological variables. However, 
ubiquitously used approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but not in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is exceedingly common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and tail dependence can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. Copula structure and tail 
dependence can also reveal causal relationships between variables. We describe consequences of copula structure, including impacts 
on extinction risk and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology.

\noindent \emph{Keywords:} dependence, regression, correlation, copula, population, community

@@

knitr::kable(tab_data_info, 
             #format="latex", align="l",
             format="latex", align="l",escape=F,
             caption = "Summary table for the data we used. Bold entries are the multivariate data, the rest are bivariate datasets (see Data section). Basal metabolic rate=BMR. \\label{tab_data_info}", 
             booktabs = T,linesep = "\\addlinespace") %>%
             column_spec(1:4,width="2.8 cm")%>%
             row_spec(c(4,5,6,8), bold=T)
```

@@

# Results for Q1: Ecological datasets often have non-normal copula structure and asymmetric tail dependence \label{Results}

@@

two_tab_in1 <- as.data.frame(do.call(cbind, lapply(1:ncol(tab1), function(i) paste0(" (",tab1[c(7:8,10:12),i], " , ",tab2[ , i], ")") )))   # see carefully if you rearrange tab_multivar_summary's row in suppmat this will change

@@

knitr::kable(tab_multivar_summary[1:9,], 
             format="latex", escape = F,
             caption.short = "Summary of results for multivariate datasets",
             caption = "Summary of Q1 results for multivariate datasets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green spruce aphid abundance",
                        "Leaf-curling plum aphid first flight",
                        "\\textit{Ceratium furca} abundance","Methane-flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

<!--**DAN: stopped here-->

Having demonstrated that non-N copula structure and asymmetric tail dependence are 
common in ecological variables, we 
addressed Q2 by exploring, using models, three possible mechanisms
by which these phenomena may arise. 
The first mechanism, summarized conceptually in the Introduction, relates to 
Liebig's law of the minimum and to nonlinear influences of environmental
variables on ecological variables.

We did not systematically 
test the first two parts of A 
because nonlinear and thresold effects of 
environmental variables on ecological variables are already well documented 
(however, see Discussion material about the green spruce aphid). To test 
the third part of hypothesis A, we formulated a model in which 
an environmental variable influences an ecological variable
more strongly in one of its tails, and explored whether the ecological variable
could then exhibit asymmetric tail dependence across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$.
The function $g$ describes how the environment $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only for negative 
values and $g_2$ only for positive values of $\epsilon$.
Negative values of $b$ correspond to overcompensating dynamics 
and positive values to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance. 
Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, the Moran 
mechanism analyzed below (Fig. \ref{MasterFigure}, B) 
does not operate here.

For each $b$ and $g_i$, we simulated the model for $25000$ time
steps and retained the $E_i(t)$ for the final $2500$ time 
steps. We applied our model selection algorithms and 
non-parametric statistics (methods for the observational hypothesis) to these 
outputs to discover if the model could
produce non-N copula structure and asymmetric tail dependence in the 
$E_i(t)$.

Hypothesis B is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
Let $E_i(t)$ again be an ecological variable, $i=1,2$.
We initially assume the AR(1) dynamics 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \epsilon_i(t)$, with $\beta=0.5$. 
The environmental noises $\epsilon_i(t)$ are 
standard-normal random variables that are independent for distinct times, $t$, but 
exhibited different kinds of 
dependence across locations in different simulations (see below).
The variable $E$ is very general, and could represent deviations
of a population density from a carrying capacity, deviations of
total plant community biomass from an average value,
flux of a biogeochemical variable such as methane, or other quantities.

To explore the importance of nonlinear dynamics for hypothesis B, we also considered 
nonlinear population models based on $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right) \right)+\sigma\epsilon_i(t))$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model into play.

For each of the model setups above, for each of several copula families, for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If causal hypothesis B is
correct, plots should be similar for populations and noises.

Hypothesis C was not tested. Our next 
causal hypothesis, Fig. \ref{MasterFigure}, D, 
is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that tail dependence occurs in evolutionary changes
in bivariate characters, and gives rise to tail dependence in observed character values
across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
bivariate draws from distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

Our model with asymmetric environmental effects 
could produce outputs with visually apparent asymmetry 
of tail dependence, to an extent that depended on the value of $b$,
thus confirming the third part of hypothesis A for our model. 
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ (Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetry of tail dependence was strong; for 
$b=\pm 0.5$, asymmetry was weaker but still apparent
(Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
Lower-tail (respectively, upper-tail) spatial dependence in 
effects of noise ($g_1$, respectively, $g_2$) 
produced lower-tail (respectively, upper-tail) spatial dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
ecological outputs had non-N copula structure, but did not always reflect 
asymmetry of tail dependence, perhaps because those methods measured
dependence in the extreme tails. Results using our nonparametric statistics, however,
reflected the visually apparent asymmetry (Tables \ref{SM-tab_stat_npa_asym}).

@@

Causal hypothesis B (Fig. \ref{MasterFigure}) was largely confirmed; 
however, similarities between environmental- and 
ecological-variable copula structure were reduced when dynamics were strongly nonlinear.
For simulations using AR(1) models, copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.
For weak noise and our nonlinear model, copula statistics
were again very similar for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-
\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}).
Though there were again significant but small differences for some statistics
and simulations, the hypothesis was supported that copula structure in the noise translates
to similar but not identical copula structure in output populations. 
Since many ecological models are nonlinear, this result provides the 
expectation that there will be a Moran-effect-like correspondence 
between noise and model-output copula structure for typical ecological dynamics,
as long as noise is small enough that dynamics stay relatively close
to the model equilibrium. Results that hold for "weak noise" in this sense
are common in ecology. 
We repeated the analysis with $r=1.3$, with similar results. The deterministic 
one-habitat-patch Ricker model exhibits a monotonic approach to a stable equilibrium
at $K$ when $r<1$ (undercompensating dynamics, e.g., the value $r=0.5$ used initially), 
but exhibits an oscillatory approach to $K$ when $r>1$ (overcompensating
dynamics, e.g., $r=1.3$).

\begin{figure}[!h]
\begin{center}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_CorlmCoru_Kendall's Tau.pdf}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_PlmPu_Kendall's Tau.pdf}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_D2umD2l_Kendall's Tau.pdf}
\caption{Example results showing similarity of copula statistics for 
environmental-noise inputs and ecological-variable outputs of a dynamical model.  
Asymmetric tail dependence was similar in both noise inputs and model outputs
for this model. The 
AR(1) model and a Clayton copula were used (see methods section for the causal hypotheses).
Each point is the mean across 50 replicate simulations for which the same 
Kendall $\tau$ value was used. Error bars are standard errors 
and panel headers give Pearson correlation results for
the points. The regression line through the points (black line) was similar
to the 1-1 line (green line). 
See Fig. \ref{SM-fig_Cause4copula_C_kend_ar1} for additional results for the AR(1) model and Clayton copula, 
Figs \ref{SM-fig_Cause4copula_SC_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1} for AR(1) results with other copula families, and 
Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1} for
the nonlinear models described in the methods for the causal hypotheses.\label{fig_Cause4copula_scatter_C_kend_ar1}}
\end{center}
\end{figure}

For strong noise and using our nonlinear model, 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to previous results, 
and for a few simulations, tail dependence asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs (Fig. \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}); 
and likewise for SG and SJ copulas (Figs \ref{SM-fig_Cause4copula_SG_kend_sRicker_nsd_1}, 
\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). Overall,
similarities between noise and model-output copula structure
were more noticable than discrepancies, in spite of nonlinearities.
We repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring. Apparently noise of standard deviation 1
interacted strongly with model nonlinearities when the model was in its overcompensatory 
regime.

For our model, hypothesis D (Fig. \ref{MasterFigure}) was 
correct that asymmetric tail dependence in evolutionary changes 
can produce similarly asymmetric tail dependence in character distributions across
phylogeny tips (Appendix \ref{SM-evol_mech}, Fig. \ref{SM-fig_cop_evo} for details). This topic 
is revisited in the Discussion.

@@

The hypothesis was presented that
the distribution (through time) of a spatially averaged quantity should be influenced by 
dependencies between the local quantities, including their
copula structure and tail dependence (Fig. \ref{MasterFigure}, X). 
To refine this hypothesis, suppose an ecological variable 
$E_i(t)$ is measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and
the spatial mean $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
right-tail dependent for most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in most locations.
We hypothesize that this 
will tend to increase the skewness of the distribution 
of the spatial mean. Similarly, left-tail dependence between local variables
should tend to decrease skewness. Strong positive skewness of the 
spatial mean corresponds to "spikiness" of the spatial-mean time series. 
The spatial-mean time series and its skewness may be quantities of 
principal importance for 
pest or resource abundance, for which extreme values (spikes) 
in the spatial mean have large effects.

We tested hypothesis X 
using our multivariate datasets (Table \ref{tab_multivar_summary}).
For each dataset we calculated the spatial mean time series, and then 
the skewness through time of that
mean. Then, for each dataset, we compared the value obtained 
to a distribution of values 
of the same quantity for each of 10000 surrogate datasets. Surrogate 
datasets were produced by randomizing the empirical data
in a special way to have the 
copula structure of a multivariate-normal distribution, but to retain exactly the same
distributions of values for each sampling location as the original data
(Appendix \ref{SM-Surrog_multi}). 
Surrogates also had very similar Spearman correlations between pairs of sampling locations
as the data. 
Our comparisons therefore tested the null hypothesis that the skewness of the spatial 
mean took values on the empirical data no different from what one would expect if the copula
structure of the data were multivariate N, but the data were otherwise statistically
unchanged. Significant differences indicate that non-N copula structure in the data 
contributed to the skewness of the spatial mean, i.e., to its "spikiness".

>>>>

\noindent \emph{Type of Submission:} Articles, or Concepts and Synthesis

@@

All branches of ecology study relationships among environmental and biological variables. However, 
standard approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but not in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is exceedingly common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and tail dependence can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. Copula structure and tail 
dependence can also help reveal causal relationships between variables. We also describe consequences of copula structure for ecological phenomena, including impacts 
on extinction risk and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology, benefiting the discipline.

\noindent \emph{Keywords:} dependence, regression, correlation, copula, population, community, ecosystem functioning

@@

knitr::kable(tab_data_info, 
             #format="latex", align="l",
             format="latex", align="l",escape=F,
             caption = "Summary table for the data we used. Bold entries are the multivariate data, the rest are bivariate datasets (see Data section). Basal metabolic rate=BMR. \\label{tab_data_info}", 
             booktabs = T,linesep = "\\addlinespace") %>%
             column_spec(1:4,width="3.4 cm")%>%
             row_spec(c(4,5,6,8), bold=T)
```

@@

# Results for Q1: Ecological datasets have non-normal copula structure and asymmetric tail dependence \label{Results}

@@

two_tab_in1 <- as.data.frame(do.call(cbind, lapply(1:ncol(tab1), function(i) paste0(" (",tab1[c(7:8,10:12),i], ", ",tab2[ , i], ")") )))   # see carefully if you rearrange tab_multivar_summary's row in suppmat this will change

@@

knitr::kable(tab_multivar_summary[1:9,], 
             format="latex", escape = F,
             caption.short = "Summary of results for multivariate datasets",
             caption = "Summary of Q1 results for multivariate datasets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green spruce aphid abundance",
                        "Leaf-curling plum aphid first flight",
                        "\\textit{Ceratium furca} abundance","Methane-flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "6cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are 
common in ecological variables, we 
addressed Q2 by exploring, using models, three possible mechanisms
by which these phenomena may arise. 
The first mechanism relates to the ideas in the Introduction about 
Liebig's law of the minimum, and to nonlinear influences of environmental
variables on ecological variables.
If an environmental variables influences an ecological variable 
more strongly in one of its tails, we explored whether the ecological
variable could then exhibit asymmetric tail dependence across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$.
The function $g$ describes how the environment $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only for negative 
values and $g_2$ only for positive values of $\epsilon$.
Negative values of $b$ correspond to overcompensating dynamics 
and positive values to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance.

For each $b$ and $g_i$, we simulated the model for $25000$ time
steps and retained the $E_i(t)$ for the final $2500$ time 
steps. We applied our model selection algorithms and 
non-parametric statistics (methods for Q1) to these 
outputs to discover if the model could
produce non-N copula structure and asymmetric tail dependence in the 
$E_i(t)$. Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, the Moran 
mechanism analyzed below 
does not operate here.

Our second mechanism is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
Let $E_i(t)$ again be an ecological variable, $i=1,2$.
We initially assume the AR(1) dynamics 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \epsilon_i(t)$, with $\beta=0.5$. 
The environmental noises $\epsilon_i(t)$ are 
standard-normal random variables that are independent for distinct times, $t$, but 
exhibited different kinds of 
dependence across locations in different simulations (see below).
The variable $E$ is very general, and could represent deviations
of a population density from a carrying capacity, deviations of
total plant community biomass from an average value,
flux of a biogeochemical variable such as methane, or other quantities.

To explore the importance of nonlinear dynamics for our proposed mechanism, 
we also considered 
nonlinear population models based on $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right)+\sigma\epsilon_i(t)\right)$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model into play.

For each of the model setups above, for each of several copula families, for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given Kendall
correlation $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were then computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If the 
hypothesis was correct that characteristics of 
the copula structure of spatial dependence in an
ecological variable can be inherited from the copula structure of spatial 
dependence in an environmental variables, then 
plots should be similar for populations and noises.

Our next mechanism is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that tail dependence occurs in evolutionary changes
in bivariate characters, and gives rise to tail dependence in the 
character values observed across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
bivariate draws from distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For 
each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

Our model with asymmetric environmental effects 
could produce outputs with visually apparent asymmetry 
of tail dependence, to an extent that depended on the value of $b$;
thus this is a mechanism that may produce non-N copula structure and
asymmetric tail dependence across space in ecological variables.
This result is represented in Fig. \ref{MasterFigure} as the solid box around 
"Nonlinear environmental effects, Liebig's law" and the arrows labelled "A".
It is explained in the Discussion why the box and some of those arrows 
are solid, instead of dashed, although the results we 
present in this section are theoretical.
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ 
(Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetry of tail dependence was strong; for 
$b=\pm 0.5$, asymmetry was weaker but still apparent
(Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
Lower-tail (respectively, upper-tail) spatial dependence in 
effects of noise ($g_1$, respectively, $g_2$) 
produced lower-tail (respectively, upper-tail) spatial 
dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
ecological outputs had non-N copula structure, but did not always reflect 
asymmetry of tail dependence, perhaps because those methods measured
dependence in the extreme tails. Results using our nonparametric statistics, however,
reflected the visually apparent asymmetry (Tables \ref{SM-tab_stat_npa_asym}).

@@

The hypothesis that characteristics of 
spatial copula structure and tail dependence
in an ecological variable can be inherited from an environmental 
variable was largely confirmed; 
however, similarities between environmental- and 
ecological-variable copula structure were reduced when 
dynamics were strongly nonlinear.
Thus this is a mechanism that may produce non-N copula structure and
asymmetric tail dependence across space in ecological variables.
This result is represented in Fig. \ref{MasterFigure} as the dashed box around 
"Moran effects" and the arrows labelled "B".
For simulations using AR(1) models, copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.
For weak noise and our nonlinear model, copula statistics
were again very similar for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-
\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}).
Though there were again significant but small differences for some statistics
and simulations, the hypothesis was supported that copula 
structure in the noise translates
to similar but not identical copula structure in output populations. 
Since many ecological models are nonlinear, this result provides the 
expectation that there will be a Moran-effect-like correspondence 
between noise and model-output copula structure across space 
for typical ecological dynamics,
as long as noise is small enough that dynamics stay relatively close
to the model equilibrium. Results that hold for "weak noise" in this sense
are common in ecology. 
We repeated the analysis with $r=1.3$, with similar results. The deterministic 
one-habitat-patch Ricker model exhibits a monotonic approach to a stable equilibrium
at $K$ when $r<1$ (undercompensating dynamics, e.g., the value $r=0.5$ used initially), 
but exhibits an oscillatory approach to $K$ when $r>1$ (overcompensating
dynamics, e.g., $r=1.3$).

\begin{figure}[!h]
\begin{center}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_CorlmCoru_Kendall's Tau.pdf}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_PlmPu_Kendall's Tau.pdf}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_D2umD2l_Kendall's Tau.pdf}
\caption{Example results showing similarity of copula statistics for 
environmental-noise inputs and ecological-variable outputs of a dynamical model.  
Asymmetric tail dependence was similar in both noise inputs and model outputs
for this model. The 
AR(1) model and a Clayton copula were used (see methods section for Q2).
Each point is the mean across 50 replicate simulations for which the same 
Kendall $\tau$ value was used. Error bars are standard errors 
and panel headers give Pearson correlation results for
the points. The regression line through the points (black line) was similar
to the 1-1 line (green line). 
See Fig. \ref{SM-fig_Cause4copula_C_kend_ar1} for additional results for the AR(1) model and Clayton copula, 
Figs \ref{SM-fig_Cause4copula_SC_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1} for AR(1) results with other copula families, and 
Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1} for
the nonlinear models described in the methods section for Q2.\label{fig_Cause4copula_scatter_C_kend_ar1}}
\end{center}
\end{figure}

For strong noise and using our nonlinear model, 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to previous results, 
and for a few simulations, tail dependence asymmetry statistics
had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs (Fig. \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}); 
and likewise for SG and SJ copulas (Figs \ref{SM-fig_Cause4copula_SG_kend_sRicker_nsd_1}, 
\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). Overall,
similarities between noise and model-output copula statistics
were more noticable than discrepancies, in spite of nonlinearities.
We repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring. Apparently 
noise of standard deviation 1 interacted especially strongly with 
model nonlinearities when the model was in its overcompensatory 
regime.

For our evolutionary model, the hypothesis was 
correct that asymmetric tail dependence in evolutionary changes 
can produce similarly asymmetric tail dependence between characters across
phylogeny tips (Appendix \ref{SM-evol_mech}, Fig. \ref{SM-fig_cop_evo} 
for details). 
Thus this is a mechanism that may produce non-N copula structure and
asymmetric tail dependence between characters.
This result is represented in Fig. \ref{MasterFigure} as the dashed box around 
"If character evolution has copula structure" and the arrow labelled "D".
This topic is revisited in the Discussion, as is the box "Asymmetric species
interactions" and the arrow labelled "C" in Fig. \ref{MasterFigure}.

@@

The hypothesis was presented in the Introduction that
the distribution (through time) of a spatially 
averaged quantity should be influenced by 
dependencies between the local quantities, including their
copula structure and tail dependence. 
To refine this hypothesis, suppose an ecological variable 
$E_i(t)$ is measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and
the spatial mean $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
right-tail dependent for most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in most locations.
We hypothesize that this 
will tend to increase the skewness of the distribution 
of the spatial mean. Similarly, left-tail dependence between local variables
should tend to decrease skewness. Strong positive skewness of the 
spatial mean corresponds to "spikiness" of the spatial-mean time series. 
The spatial-mean time series and its skewness may be quantities of 
principal importance for 
pest or resource abundance, for which extreme values (spikes) 
in the spatial mean have large effects.

To help answer Q3, we tested the above hypothesis
using our multivariate datasets (Table \ref{tab_multivar_summary}).
For each dataset we calculated the spatial mean time series, and then 
the skewness through time of that
mean. Then, for each dataset, we compared the value obtained 
to a distribution of values 
of the same quantity for each of 10000 surrogate datasets. Surrogate 
datasets were produced by randomizing the empirical data
in a special way to have the 
copula structure of a multivariate-normal distribution, but 
to retain exactly the same
distributions of values for each sampling location as the original data
(Appendix \ref{SM-Surrog_multi}). 
Surrogates also had very similar Spearman correlations 
between pairs of sampling locations
as the data. 
Our comparisons therefore tested the null 
hypothesis that the skewness of the spatial 
mean took values on the empirical data no different 
from what one would expect if the copula
structure of the data were multivariate N, but the data were otherwise statistically
unchanged. Significant differences indicate that non-N copula structure in the data 
contributed to the skewness of the spatial mean time series, 
i.e., to its "spikiness".
